---
title: "Data Vis Project Portugese"
author: "Levon Yesayan"
date: "12/1/2022"
output:
  pdf_document:
    number_sections: TRUE
    latex_engine: xelatex
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = F, message = F)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(ggExtra)
library(readxl)
library(viridis)
library(tidyverse)
library(scales)
library(ggpubr)
library(devtools)
library(ggthemes)
```


```{r cars}
data_por <- read.csv('student-por.csv', sep=";")
data_por <- transform(data_por, avrg_G=(G1+G2+G3)/3)
head(data_por)
```
```{r cars1}

data_cor<-data[c("G1","G2","G3","absences","Walc","Dalc","health","goout","freetime","famrel","failures","studytime","traveltime","age")]

corr_mat <- round(cor(data_cor),2)
melted_corr_mat <- melt(corr_mat)

ggplot(data = melted_corr_mat, aes(x=X1, y=X2, fill=value)) + geom_tile() + geom_text(aes(X2, X1, label = value), color = "black", size = 4) + theme(axis.text.x = element_text(angle = 90))


```
Based on very strong correlation between G1, G2 and G3, which correspond to first, second periods and final grade respectively, we can merge them into one average grade and continue the study considering average_grade as the dependent variable.


Hypothesis 1: Students in urban areas perform better than students in rural areas.

```{r cars2}
address.labs <- c("Rural", "Urban")
male_female <- c("Female", "Male")
names(address.labs) <- c("R", "U")
ggplot(data_por, aes(x=sex, y=avrg_G, fill=sex)) + 
  geom_boxplot(notch=TRUE,colour = "#020427") + 
  facet_grid(~address, labeller = labeller(address = address.labs))  + 
  labs(title = "Student Performance by geographical location and gender",
       y = "Average grade") +
  scale_x_discrete(labels= male_female) + theme_stata() +
  scale_fill_stata(name = "Gender", labels = c("Female", "Male")) +
  theme(axis.title.x = element_blank()) 
```

The graph shows that students who live in an urban location have higher performance than those who live in rural areas. A statement can be 
made that in both geographical regions, female students perform better than male students on average. Since the notches in the box plot do not overlap, we can conclude, with 95% confidence, that the actual medians do differ. Finally, we do not reject our null hypothesis stating that urban students perform better than rural students.

Hypothesis 2: The quality of the family relationship has positive effect on student's performance.

```{r cars3}
quality_rel <- c("Very bad", "Bad","Average","Good","Extremely good")
ggplot(data_por, aes(x=as.character(famrel), y=avrg_G, color = famrel)) + 
  geom_jitter(position=position_jitter(0.2)) +
  labs(title = "Student Performance Based on Quality of Relationship with Family",
       y="Average Grade", x="Quality of Relationship with Family") +
  scale_x_discrete(labels= quality_rel)  + scale_color_viridis() + theme_clean()+
  theme(legend.position = "none") + stat_summary(fun = mean, geom = "point", 
                                                 shape = 18,size=4,color="red",
                                                 fill = "red")

```
The plot implies that most students have average to excellent relationships with their families. Moreover, their performance does not guarantee that they correlate with the relationship quality with family, as the observations are almost equally distributed. Thus we reject our null hypothesis, which states that the quality of the family relationship has a positive effect on student performance.

Hypothesis 3: Parent's education has positive effect on student's performance.

```{r cars4}
table(data_por$Medu)
table(data_por$Fedu)
```

```{r cars4.1}
Medu.labs <- c("Primary education","5th to 9th grade"
               ,"Secondary education ","Higher education")
names(Medu.labs) <- c("1","2","3","4")
Fedu.labs <- c("Primary education","5th to 9th grade"
               ,"Secondary education ","Higher education")
names(Fedu.labs) <- c("1","2","3","4")

means1_por <- aggregate(avrg_G ~ Medu, data = data_por[data_por['Medu'] != "0",], FUN=mean)
means2_por <- aggregate(avrg_G ~ Fedu, data = data_por[data_por['Fedu'] != "0",], FUN=mean)

pe_por<-ggplot(data_por[data_por["Medu"] != "0",], aes(x = avrg_G,fill= Medu)) + geom_histogram(bins = 25,alpha = 0.8) + 
  labs(x = "Average Grade", y = "Frequency",
       title = "Student’s performance based on Mothers’s education") +
  facet_grid( ~ Medu, scales = "free",labeller = labeller(Medu = Medu.labs)) +
  scale_fill_viridis(option = "plasma") + theme_hc() +
  theme(legend.position = "none") + scale_x_continuous(limits=c(0,20)) +
  geom_vline(data =means1_por, aes(xintercept=avrg_G), colour="red") 

pe2_por<-ggplot(data_por[data_por["Fedu"] != "0",], aes(x = avrg_G,fill= Fedu)) + geom_histogram(bins = 25,alpha = 0.8) + 
  labs(x = "Average Grade", y = "Frequency",
       title = "Student’s performance based on Father’s education") +
  facet_grid( ~ Fedu, scales = "free",labeller = labeller(Fedu = Fedu.labs)) +
  scale_fill_viridis(option = "plasma") + theme_hc() +
  theme(legend.position = "none") + scale_x_continuous(limits=c(0,20)) +
  geom_vline(data =means2_por, aes(xintercept=avrg_G), colour="red") 

figure_pe_por <- ggarrange(pe_por, pe2_por, ncol = 1, nrow = 2)
figure_pe_por
```


Firstly the figure implies that there nor much evidence of parents not having any education. The most common grades for students whose parents have primary education varies around 10, which is a bad result overall. Common grades for students whose parents have 5th to 9th-grade education vary between 10 to 12, which is a bad result again. There is a similar pattern for students whose parents have only secondary compared to the previous group.Finally, most common grades for students whose parents have higher education varies around 14-16, which is a decent result  overall. In conclusion, we do not reject the null hypothesis, which states that parents' education  positively affect student performance. However, parents with at least 5th to 9th-grade education already affect their child's academic performance positively, and there is small gradual increase in performance for parents with higher educations.

Hypothesis 4: Student's with intention of applying for higher education have significantly higher average grade.

```{r cars5}
##ggthemes("light")
ggplot(data_por,aes(x = higher, y =avrg_G , fill = higher)) + 
  geom_boxplot() + geom_dotplot(binaxis = "y", stackdir = "center", dotsize = 0.5)+
  labs(title = "Peroformace of students by ambition for higher education",
       x = "Inclination for higher education",y = "Average Grade",fill="")
```
First, the data provide insights that most students have the ambition of higher education. Still, compared with students who are not willing to apply for higher education, their performance is better overall. Thus, we do not reject our null hypothesis stating that students who use higher education have significantly higher average grades than those not willing to continue higher education.

Hypothesis 5: Different kinds of educational support boost the performance of the student significantly.

```{r cars6}

data_por$concatenated<-paste(data_por$schoolsup,data_por$famsup,data_por$paid)
agg_concat_por <- aggregate(avrg_G~concatenated,data =data_por, FUN = mean)
ggplot(data_por, aes(x=concatenated, y=round(avrg_G,2), fill=concatenated)) +
  geom_bar(stat="identity", position = 'dodge') + 
  labs(fill = "EES/ FES /EPC ", 
       title = "Student performance based in exta educational support",
       subtitle = "(Extra Educational Support / Family Educational Suport / Extra Paid Classes)",
       y = "Average Grade", x = "Combination of educational help") +
  theme(axis.text.x = element_blank())
```
Hypothesis 5: Intense alcohol consumption strongly decreases students’ performance based on gender

```{r cars7}
agg_data <- aggregate(avrg_G ~ Dalc, data = data_por, FUN=mean)
ggplot(agg_data, aes(x = Dalc, y = round(avrg_G,2))) + geom_line(size=2, color='orange') + geom_point(color = 'red', size=3) + labs(x = 'Daily Alcohol Consumption', y = "Average Grade of Exams") + geom_text(aes(label=round(avrg_G,2)), position=position_dodge(width=0.9), vjust=-1, size = 4) + theme() + scale_fill_manual(values=c('salmon','darkturquoise'), labels=c('Female', 'Male')) + guides(fill=guide_legend("Gender")) + theme_bw() + ggtitle('Average grade of students based on daily alcohol consumption') + ylim(8,12.5) + scale_x_continuous(breaks=c(1,2,3,4,5), labels=c("Very low","Low","Medium","High", "Very high"))

```
Hypothesis 6: Being in a romantic relationships harms students’ average grade based on gender.

```{r cars8}
agg_data <- aggregate(avrg_G ~ romantic + sex, data = data_por, FUN=mean)
ggplot(agg_data, aes(x = romantic, y = round(avrg_G,2), fill = sex)) + geom_bar(stat='identity', position = 'dodge') + labs(x = 'In a romantic relationship', y = "Average Grade of Exams") + geom_text(aes(label=round(avrg_G,2)), position=position_dodge(width=0.9), vjust=-0.3, size = 4) + scale_fill_manual(values=c('salmon','darkturquoise'), labels=c('Female', 'Male')) + guides(fill=guide_legend("Gender")) + theme_few() + ggtitle('Average grade of students based on romantic relationship') + ylim(0,12.5) + theme(panel.background = element_rect(fill = "#FFC4C47D")) +  scale_x_discrete(breaks=c("no","yes"), labels=c("No", "Yes"))
  
```

Hypothesis 7: It is significantly better to live alone as a student in terms of academic performance.

```{r cars9}

summarised <- data_por %>% group_by(Pstatus) %>% summarise(mean = mean(avrg_G), sd = sd(avrg_G))
ggplot(summarised, aes(x = Pstatus, y = mean, ymin=mean-sd, ymax = mean+sd)) + geom_errorbar(size=0.5) + geom_point(colour = 'Red', size = 4) + labs(x = 'Cohabitation status', y = "Expected grade CI") + ggtitle("Confidence interval of the expected grade based on cohabitation with parents") +  scale_x_discrete(breaks=c("A","T"), labels=c("Apart", "Together")) + theme_hc()

data_por$Pstatus2 <- data_por$Pstatus

data_por["Pstatus2"][data_por["Pstatus2"]=='A']<-0
data_por["Pstatus2"][data_por["Pstatus2"]=='T']<-1

anova <- aov(Pstatus2 ~ avrg_G , data = data_por)

summary(anova)


```

Hypothesis 8: Students who do not hang out with friends are expected to have better performance.

```{r cars10}


agg_data <- aggregate(avrg_G ~ goout, data = data_por, FUN=mean)

agg_data[2,2] <- (agg_data[2,2] + agg_data[3,2])/2
agg_data[3,2] <- (agg_data[4,2] + agg_data[5,2])/2
agg_data<- agg_data[c(-4,-5),]

ggplot(agg_data, aes(x = goout, y = round(avrg_G,2))) + geom_segment(aes(x=goout, xend=goout, y=0, yend = avrg_G), size=1) + geom_point(size=5, color="red", fill=alpha("orange"), alpha=1, shape=21, stroke=2) + labs(x = 'Hangout with friends', y = "Average Grade of Exams") + geom_text(aes(label=round(avrg_G,2)), position=position_dodge(width=0.9), vjust=-1.2, size = 4) + scale_fill_brewer(palette = 'Set2', labels=c('Little','Medium', 'High')) + guides(fill=guide_legend("Hangout")) + scale_x_discrete(limits=c("1","2","3"),labels = c("Low","Medium","High")) + theme_classic() + ggtitle('Average grade of students based on student hangout') + ylim(0,13) 


```
Hypothesis 9: Many absences lead to many failures.

```{r cars11}
correlation <- cor(data_por$failures, data_por$absences)

ggplot(data_por, aes(x = failures, y = absences)) + geom_point() + geom_jitter(height = 0.1, width = 0.1) + geom_smooth(formula=y~x,method = 'lm', se=FALSE) + theme_classic() + labs(x = 'Failures', y = "Absences", caption = paste0('Correlation coefficient ', round(correlation,3))) + ggtitle('Student absences vs failures')

```

Hypothesis 10: Students studying less are expected to have significantly lower average grade. 

```{r cars11}

ggplot(data = data_por, aes(x = studytime, y=avrg_G, group = studytime, fill = factor(studytime))) + geom_violin() + geom_boxplot(width = 0.3) + ggtitle('The distribution boxplot of average grade based on study time of the students') + labs(x='Study hours (weekly)', y = 'Average grade') + theme_hc() +  theme(legend.position = "none") + scale_fill_brewer(palette='Pastel2') + scale_x_discrete(limits=c("1","2","3","4"),labels = c("<2","2-5","5-10",">10"))

```



